---
title: "chirps-stars"
output: github_document
bibliography: D:/MendeleyBib/Z_GIS.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Exploration of CHIRPS data using R and the stars package for handling NetCDF files.*

## CHIRPS data

"The Climate Hazards Group Infrared Precipitation with Station data (CHIRPS) is a high-resolution climatic database of precipitation embracing monthly precipitation climatology, quasi-global geostationary thermal infrared satellite observations from the Tropical Rainfall Measuring Mission (TRMM) 3B42 product, atmospheric model rainfall fields from National Oceanic and Atmospheric Administration – Climate Forecast System (NOAA CFS), and precipitation observations from various sources." [@Retalis2017]

Overall it is a combination of station-base observations around the world, and satellite derived data for meteorology, including TRMM and NOAA, which are normally used for precipitation analysis. 

The data covers information from 1981 to the present on a daily, monthly, yearly and decadal basis. It spans latitudes between 50°S and 50°N and all longitudes. The data is presented in a raster format with a cell siye of 0.05° or 0.25°. Each of the cells represent a rain gauge [@Funk2015]. 

## Data download

The global daily data which will be analyzed can be downloaded per year, either as a NetCDF file, where each band corresponds to a day; or daily as TIFF files. In this case I decided to work with NetCDF files and will explain how to explore the data using R and the `stars`package. 

To explore initially the data, I will use the year 2019 at 0.05° resolution, which has information up to February. The updated data can be downloaded from this link (ftp://ftp.chg.ucsb.edu/pub/org/chg/products/CHIRPS-2.0/global_daily/netcdf/p05/)[^1]. 

[^1]: ftp links not working in Markdown.

## Loading data into R

The focus of this data exploration is to use NetCDF files in R. This file format is meant for array oriented storage. It can have multiple dimesions and its applications span climatology, meteorology, oceanography, and general GIS data handling. 

Packages such as `netcdf4` and `RNetCDF` are meant to handle such data in R. However, the new `stars` package is a promising approach to integrate NetCDF files with existing spatial packages for R, i.e. `sf`. 

```{r, echo = T, eval = F}
devtools::install_github('r-spatial/stars')
```

To load the data, `stars` makes use of its generic `read_stars` function which allows reading raster/array data. The function `read_ncdf` is also available and uses the netcdf package directly to load. 

```{r}
library(stars)
chirps19 <- read_stars('chirps-v2.0.2019.days_p05.nc')
chirps19
```

Setting a projection:

```{r}
chirps19 <- st_set_crs(chirps19, value = 4326)
chirps19
```





# References